---
title: "Hybrid Imputation Models Through Blocks"
author: "Stef van Buuren"
date: "2018/8/1, JSM 2018, Vancouver"
output: 
  beamer_presentation:
    theme: "default"
    colortheme: "beaver"
    keep_tex: true
---

# Imputing multivariate missing data

Strategies

- Joint model (JM)
- Fully conditional specification (FCS)

Questions

- Can these be complementary strategies? 
- When could hybrid imputation be useful?

# Joint model (JM)

- Specify joint model $P(Y, X)$
- Derive $P(Y^\mathrm{mis} | Y^\mathrm{obs}, X)$
- Draw imputations $\dot Y^\mathrm{mis}$ with Gibbs sampler

# JM 

Pro

- Conditionals are compatible
- Yield correct statistical inference under assumed JM
- Efficient parametrization possible
- Known theoretical properties

Con

- Lack of flexibility
- Can assume more than the complete data problem 
- Leads to unrealistically large models

# Fully conditional specification (FCS)

- Specify $P(Y^\mathrm{mis} | Y^\mathrm{obs}, X)$
- Draw imputations $\dot Y^\mathrm{mis}$ with Gibbs sampler

# FCS example: Multivariate Imputation by Chained Equations (MICE)

- Specify imputation models $P(Y_j^\mathrm{mis} | Y_j^\mathrm{obs}, Y_{-j}, X)$
- Fill in starting imputations
- And iterate

# Fully conditional specification (FCS)

Pro 

- Extremely flexible, close to the data
- Subset selection of predictors
- Modular, can preserve valuable work
- Appears to work very well in practice
- Easy to explain

Con 

- Theoretical properties only known in special cases
- Possible incompatibility
- No computational shortcuts

# Hybrids of JM and FCS

- Partition variables into $b$ *blocks* $h = 1,\dots,b$
- Example:

b  |  partioning                          | model
---|--------------------------------------|--------
4  | $\{Y_1\},\{Y_2\},\{Y_3\},\{Y_4\}, X$ | MICE
2  | $\{Y_1, Y_2, Y_3\},\{Y_4\}, X$       | hybrid
1  | $\{Y_1, Y_2, Y_3, Y_4\}, X$          | JM

# JM embedded within FCS

b  | h | target            | predictors| type
---|---|-------------------|-----------|------
2  | 1 | $\{Y_1, Y_2, Y_3\}$ | $Y_4, X$ | mult
2  | 2 | $Y_4$ | $Y_1, Y_2, Y_3, X$     | univ

# FCS embedded within FCS 

b  | h | j | target| predictors      | type
---|---|---|-------|------------------|--------
2  | 1 | 1 | $Y_1$ | $Y_2, Y_3, Y_4, X$ | univ
2  | 1 | 2 | $Y_2$ | $Y_1, Y_3, Y_4, X$ | univ
2  | 1 | 3 | $Y_3$ | $Y_1, Y_2, Y_4, X$ | univ
2  | 2 | 1 | $Y_4$ | $Y_1, Y_2, Y_3, X$ | univ

# Consequence of blocks in `mice()`

- Main iteration loop over `blocks`
- `dim(predictorMatrix)`: $b \times p$
- `length(methods)`: $b$ (instead of $p$)

# Example: Multilevel data

```{r mldata}
library(mice, warn.conflicts = FALSE)
library(miceadds)
d <- brandsma[, c("sch", "lpo", "iqv", "den")]
```

# FCS multilevel (c.f. Chapter 7, FIMD2)

```{r mlfcs, eval = FALSE}
meth <- make.method(d)
meth[c("lpo", "iqv", "den")] <- c("2l.pmm", "2l.pmm",
                                  "2lonly.pmm")
pred <- make.predictorMatrix(d)
pred["lpo", ] <- c(-2, 0, 3, 1)
pred["iqv", ] <- c(-2, 3, 0, 1)
pred["den", ] <- c(-2, 1, 1, 0)
imp <- mice(d, pred = pred, meth = meth, seed = 418,
            m = 10, print = FALSE)
```

# Idea 1: Embed joint model for multilevel data

```{r jomoImpute, eval = FALSE}
# mitml::jomoImpute / called from mice
d$den <- as.factor(d$den)
blk <- make.blocks(d, "collect")
fm <- list(collect = list(lpo + iqv ~ 1 + (1 | sch), 
                          den ~ 1))
imp <- mice(d, meth = "jomoImpute", blocks = blk, 
            form = fm, print = FALSE, seed = 1, 
            maxit = 2, m = 10, n.burn = 100)
```


# Idea 2: Model-based imputation

- Define complete-data model $P(Y_1 | Y_2, X)$
- Specify imputation model $P(Y_2 | X)$ (no $Y_1$!)
- Rejection sampling to target $P(Y_2 | Y_1, X)$
- Wu 2010, Bartlett 2015, Erler 2016
- `R` Software: `smcfcs`, `mdmb`, `jomo`
- Imputation compatible with complete-data model
- Can preserve deterministic relations
- Useful for strong, pre-specified complete-data models


# Idea 3: Multivariate predictive mean matching

- Idea: Impute vector instead of scalar value
- Use redundancy analysis (van den Wollenberg 1977)
- Predictive mean matching using redundancy predictor
- Expected result: consistency within blocks, faster convergence

# Idea 4: Imputing measurement scales

- Idea: Create module for imputing items + sum score
- Scale: items + sum score; some items are missing
- Items predicted by 
    1) other items in scale 
    2) sum scores other scales
- Expected result: simplification of model specification

# `predictorMatrix` simplification: current method

```{r echo=FALSE}
nam <- c("age", "a1", "a2", "s_a", "b1", "b2", "b3", "s_b")
a <- matrix(c(0, 0, 0, 1, 0, 0, 0, 1,
              1, 0, 1, 0, 0, 0, 0, 1,
              1, 1, 0, 0, 0, 0, 0, 1,
              0, 1, 1, 0, 0, 0, 0, 0,
              1, 0, 0, 1, 0, 1, 1, 0,
              1, 0, 0, 1, 1, 0, 1, 0,
              1, 0, 0, 1, 1, 1, 0, 0,
              0, 0, 0, 0, 1, 1, 1, 0),
              byrow = TRUE, nrow = 8, 
            dimnames = list(nam, nam))
print(a)
```

# `predictorMatrix` simplification: new method

```{r}
blocks <- list(age = "age", 
               A = c("a1", "a2", "s_a"), 
               B = c("b1", "b2", "b3", "s_b"))
```
```{r echo = FALSE}
namr <- c("age", "A", "B")
b <- matrix(c(0, 0, 0, 1, 0, 0, 0, 1,
              1, 0, 0, 0, 0, 0, 0, 1,
              1, 0, 0, 1, 0, 0, 0, 0),
              byrow = TRUE, nrow = 3, 
            dimnames = list(namr, nam))
print(b)
```

# Idea 5: Combine imputation models from overlapping data

Source 1: $\{Y_1, Y_2\}$ given $X$ - prefitted

Source 2: $\{Y_1, Y_3\}$ given $X$ - prefitted

b  | h | target            | predictors| type
---|---|-------------------|-----------|------
2  | 1 | $\{Y_1, Y_2\}$ | $Y_3, X$ | mult
2  | 2 | $\{Y_1, Y_3\}$ | $Y_2, X$ | mult

# Conclusion

- Blocks are conceptually straightforward extension
- `blocks` implemented in `mice 3.0`
- Documentation still in the works
- Easy to specify hybrids of JM and FCS
    - Idea 1: Embed joint model for multilevel data
    - Idea 2: Model-based imputation
    - Idea 3: Multivariate predictive mean matching
    - Idea 4: Imputing measurement scales
    - Idea 5: Combine imputation models


